Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-05-28 main.Rmd
number of observation write-up done

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $N_C \choose 2$. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left could be large and overwhelming for human consumption. Hence, there is a need to rank these harmonies basis how well they capture the variation in the measured variable. The harmony pairs $(C_i, C_j)$ with higher average variation in the measured variable between different levels of $C_i$ and $C_j$ has a higher chance to bring out distinct characteristics of the data.<!-- Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. -->

@@

\noindent One of the potential ways to evaluate this variation is by computing the pairwise distances between the distributions of the measured variable. <!--across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. ---> We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence which is based on Kullback-Leibler divergence. Probability distributions are represented through sample quantiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. The Jensen-Shanon distance between two sample quantile vectors $q_u$ and $q_v$ computed for probability vector $p$ will be given by $$d_{uv} = [D(q_u, r) + D(q_v, r)]/2 \quad where \quad r = (q_u + q_v)/2$$ where,
 $$D(q_u,q_v) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$ is the Kullback-Leibler divergence between $q_u$ and $q_v$.

@@

\noindent The harmony pairs could be arranged from highest to lowest average maximum pairwise distances across different levels of the cyclic granularities. But maximum is not robust to the number of levels and seems to be higher for harmonies with higher levels. Thus these maximum pairwise distances need to be normalized for different harmonies in a way that eliminates the effect of different levels. \noindent The Fisher–Tippett–Gnedenko theorem in the field of Extreme Value Theory states that the maximum of a sample of iid random variables after proper re normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process. Suppose $X_{1},X_{2}\ldots ,X_{n}$  be a sequence of independent and identically-distributed random variables, and $M_{n}=\max\{X_{1},\ldots ,X_{n}\}$. If a sequence of pairs of real numbers $a_{n}>0$ and $\lim _{{n\to \infty }}P\left({\frac  {M_{n}-b_{n}}{a_{n}}}\leq x\right)=F(x)$, where $F$ is a non-degenerate distribution function, then the limit distribution $F$ belongs to either the Gumbel, the Fréchet or the Weibull family. The normalizing constants, however, vary depending on the underlying distribution of $X_i$'s. Hence to normalize appropriately, it is important to assume a distribution of distances in our case.

@@

\noindent The algorithm employed addressing all the issues decried above is summarized as follows:
<!-- Algorithm -->

@@

<!-- T+1 observation not T -->

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities is a valid way to compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

However, we need to consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ and $C_2$ with $l$ and $m$ categories respectively. Each element of $C_1$ occurs approximately $T/l$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(lm)$. If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge lmk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (\ref{sec:summarize-measured}) or unevenly distributed events.

@@

# filter(fielding_wckts %in% c(0,1)) %>%
#
 cricket_data %>%
  filter(over!=1) %>%
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "boxplot",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)) +
   #ggtitle("(b) Runs per over across overs faceted by number of wickets in previous over") +
  ylab("runs per over")  +
  xlab("number of wickets in previous over") +
  ggtitle("b") +
   theme(plot.title = element_text(face = "bold")) +  theme_minimal()
# + geom_boxplot(width = 0.5, aes(colour = lag_field))
```

>>>>

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $N_C \choose 2$. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left could be large and overwhelming for human consumption. Hence, there is a need to rank these harmonies basis how well they capture the variation in the measured variable.<!--The harmony pairs $(C_i, C_j)$ with higher average variation in the measured variable between different levels of $C_i$ and $C_j$ has a higher chance to bring out distinct characteristics of the data.<!-- Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. -->

@@

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable across different levels of the cyclic granularities. <!--across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. ---> We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence which is based on Kullback-Leibler divergence. Probability distributions are represented through sample quantiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. The Jensen-Shanon distance between two sample quantile vectors $q_u$ and $q_v$ computed for probability vector $p$ will be given by $$d_{uv} = [D(q_u, r) + D(q_v, r)]/2 \quad where \quad r = (q_u + q_v)/2$$ where,
 $$D(q_u,q_v) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$ is the Kullback-Leibler divergence between $q_u$ and $q_v$.

@@

\noindent The harmony pairs could be arranged from highest to lowest average maximum pairwise distances across different levels of the cyclic granularities. But maximum is not robust to the number of levels and seems to be higher for harmonies with higher levels. Thus these maximum pairwise distances need to be normalized for different harmonies in a way that eliminates the effect of different levels. \noindent The Fisher–Tippett–Gnedenko theorem in the field of Extreme Value Theory states that the maximum of a sample of iid random variables after proper re-normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process. Suppose $X_{1},X_{2}\ldots ,X_{n}$  be a sequence of independent and identically-distributed random variables, and $M_{n}=\max\{X_{1},\ldots ,X_{n}\}$. If a sequence of pairs of real numbers $a_{n}>0$ and $\lim _{{n\to \infty }}P\left({\frac  {M_{n}-b_{n}}{a_{n}}}\leq x\right)=F(x)$, where $F$ is a non-degenerate distribution function, then the limit distribution $F$ belongs to either the Gumbel, the Fréchet or the Weibull family. The normalizing constants, however, vary depending on the underlying distribution of $X_i$'s. Hence to normalize appropriately, it is important to assume a distribution of distances in our case.

@@

\noindent The algorithm employed addressing all the issues described above is summarized as follows:
<!-- Algorithm -->

@@

<!-- For a tsibble with index ranging from ${1, 2, \dots, (T + 1)}$, that is (T+1) observations -->

<!-- In linear granularities, if we consider a data set of size T where each time point corresponds to an unique value of the measured variable.  -->

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. <!--This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable.--> It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of the measured variables through these multiple observations across cyclic granularities could be a potential way to explore and bring forward different features of the  data. <!-- compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

<!-- using some summary statistics, each of which can highlight different features of the data or through probability densities -->

<!-- T+1 observation not T -->
However, we need to consider the effect of number of observations on the summarization even for harmonies. Consider a data set with $(T + 1)$ observations, and two cyclic granularities $C_i$ and $C_j$ with $L$ and $M$ categories respectively <!--Each element of $C_i$ occurs approximately $(T+1)/L$ times while each element of $C_j$ occurs approximately $(T+1)/M$ times. and there are no or very few empty combinations since we only consider harmonies.--> Each combination might occur on average an equal number of times as $T\rightarrow\infty$ when the number of observations per combination is $nobs = (T + 1)/(LM)$ or they can vary anywhere between $nobs = 0$ and $nobs = (T+1)$. Now the statistical transformation used for summarizing the measured variable  for a combination should be chosen in line with this $nobs$. For example, it might be useful to compute deciles only when $nobs>=10$. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (\ref{sec:summarize-measured}) or unevenly distributed events. For Kernel density based plots $nobs>30$ is suggested, whereas for summary statistics based plots like boxplots $nobs = 10$ is acceptable and for quantile plots $nobs$ should at least be equal to the number of quantiles plotted.

<!-- If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge lmk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested.  -->

@@

# filter(fielding_wckts %in% c(0,1)) %>%
#
 cricket_data %>%
  filter(over!=1) %>%
  prob_plot("over", "lag_field",
  hierarchy_model,
  response = "run_rate",
  plot_type = "quantile",
  symmetric = FALSE,
 quantile_prob = c(0.25, 0.5, 0.75)) +
   #ggtitle("(b) Runs per over across overs faceted by number of wickets in previous over") +
  ylab("runs per over")  +
  xlab("number of wickets in previous over") +
  ggtitle("b") +
   theme(plot.title = element_text(face = "bold")) +  theme_minimal()
# + geom_boxplot(width = 0.5, aes(colour = lag_field))
```
